\documentclass{article}
\usepackage{amsmath}

\title{A Paper with Weak Logical Structure}
\author{Bob Tester}

\begin{document}
\maketitle

\begin{abstract}
We hypothesize that reinforcement learning can solve all NP-hard optimization problems
in polynomial time. We test our approach on a simple sorting task and achieve perfect accuracy.
Our results demonstrate the superiority of RL over all classical algorithms.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Classical algorithms struggle with NP-hard problems. We propose using deep reinforcement
learning to solve arbitrary NP-hard optimization problems efficiently.

Our key hypothesis is that a sufficiently large neural network trained with RL
can approximate optimal solutions to NP-hard problems in polynomial time.

\section{Methodology}
\label{sec:method}

We train a small MLP with 2 hidden layers using Q-learning on a sorting task
with lists of 10 integers. We use a reward of +1 for correct sorts and -1 otherwise.

\section{Results}
\label{sec:results}

Our method achieves 100\% accuracy on sorting lists of 10 integers.
Training converges in 500 episodes. However, accuracy drops to 45\% on lists of 100 integers.

\section{Conclusion}
\label{sec:conclusion}

We have proven that reinforcement learning can solve NP-hard problems in polynomial time.
Our revolutionary approach makes classical algorithms obsolete and will transform
computer science. All optimization problems are now tractable with our method.

\begin{thebibliography}{9}
\bibitem{sutton2018} Sutton, R. and Barto, A. (2018). Reinforcement Learning: An Introduction. MIT Press.
\end{thebibliography}

\end{document}
